---
title: "Question_5"
author: "Mashhood Syed"
date: "October 2, 2014"
output: pdf_document
---

  Dataset 1 is based on US Census data and was part of a study done by some 
  researchers at UCI.  The dataset was used to in a ML algorithm to determine 
  whether or not someone made >= $50k/year.  
  The link to the dataset is:  http://archive.ics.uci.edu/ml/machine-learning
  -databases/adult/.

  Dataset 2 is the count of people that came through 1 of three Express checkout 
  cafe registers at the Wegman's in Hunt Valley on Friday October 3rd, 2014 between   
  3:07pm and 5:07pm.  The data has 3 variables:  Count of people, Hour 
  Interval (3:07-4:07 = 1, 4:07-5:07 = 2), and register number.
  
  Dataset 3 is the count of people that walked into Starbucks Coffee on Queen St in 
  York, PA on Sunday, October 5, 2014 between 12pm and 2pm.  The data has 3   
  variables:  Count of people, 5 minute interval, and hour interval.
  
Question 5 Part 1
Dataset 1:
    We first set up the working directory, grab a subset of the original dataset
    and rename columns.
```{r}
  setwd("~/Desktop/Problem Set 1")
  ds1 = "~/Desktop/Problem Set 1/random_variable1.tsv"
  data1 = read.csv(ds1, header = F, sep = ",")
  df1 = subset(data1, select = c("V1","V4","V6","V7","V9","V10","V15"))
  colnames(df1)[1] = "Age"
  colnames(df1)[2] = "Education"
  colnames(df1)[3] = "Marital-status"
  colnames(df1)[4] = "Occupation"
  colnames(df1)[5] = "Race"
  colnames(df1)[6] = "Gender"
  colnames(df1)[7] = "IncomeLevel"
```
  
    Here's a histogram representing the PMF for the Age Variable
```{r, echo=FALSE}
hist(df1$Age, prob=T, ylim=c(0, 0.08), breaks = 200, main = "Age (pmf)", xlab = "Age", las = 1)
lines(density(df1$Age), col=2, lwd=3)
```
    Heres another histogram representing the frequency distribution for the Age    
    Variable (using ggplot2)
```{r}
library(ggplot2)
x = ggplot(df1, aes(Age))
x + geom_histogram()
```
    Here is a stacked histogram (using ggplot2) showing the distrbution of Sex for   
    each Race
```{r}
x = ggplot(df1, aes(Race, fill = Gender) )
x + geom_histogram()
```

    Here is a simple CDF for the Age Variable
```{r}
p = ecdf(df1$Age)
cdfAge = plot(p, ylab="Fn(Age)", xlab="Age", log="x", main="Age (cdf)")
```

    Here is a more sophisticated CDF for the Age Variable
```{r}
library(ggplot2)
ggplot(df1, aes(x = df1$Age),) + stat_bin(aes(y = cumsum(..count..)), geom = "step", colour = "black", fill = "white", binwidth = 5) + theme_bw()
```

Question 5 Part 2.
Dataset 1:

  Here is a summary of the data set
```{r}
summary(df1)
```
  Here is a summary of the Age variable
```{r}
summary(df1$Age)
```
  Here is a boxplot of the Age variable
```{r}
boxplot(df1$Age, main="Boxplot of Age", ylab="Age", ylim=c(0,100))
```
  Here is a boxplot using ggplot2 with Data Points for the Age variable
```{r}
ggplot(df1, aes(x = factor(0), y = Age)) + geom_boxplot(width = .2) + ylab("Age") + xlab("") + geom_point(aes(x = factor(0), y = Age), size = 1, colour = "blue") + theme_bw()
```

Question 5 Part 3
Dataset 1: 

    Here is the MLE for the Age Variable
```{r}
theta = c(0,1)
fn = function(theta) { sum( 0.5*(df1$Age-theta[1])^2/theta[2] + 0.5* log(theta[2]) ) }
nlm(fn, c(0,2), hessian = TRUE)
```
  I am not that sure on this part but from what I read online it looks as if the
  estimate of 38.58162 is very close to my actual mean of: 
```{r}
mean(df1$Age)
```
Question 5 Part 4
Dataset 1:

  I expected to see a normal distribution.  What I found is a normal distribution     
  with a "long right tail" in the 3rd and 4th quartiles.  I think the reason for 
  this is because there just are not that many people that still work for an income 
  once you get into the age range of 65+.

#----------------------------------------------------------------------------------

Question 5 Part 1
Dataset 2:
    We first set up the working directory and place the data in a dataframe.
```{r}
  setwd("~/Desktop/Problem Set 1")
  ds2 = "~/Desktop/Problem Set 1/random_variable2.tsv"
  df2 = read.csv(ds2, header = T, sep = ",")
```
  
    Here's a histogram representing the PMF for the Count of People:
```{r, echo=FALSE}
hist(df2$PPLCount, prob=T, ylim=c(0, 0.10), breaks = 7, main = "Count of People (pmf)", xlab = "Count", las = 1)
lines(density(df2$PPLCount), col=2, lwd=3)
```
    Heres another histogram representing the frequency distribution for the PPLCount    
    Variable (using ggplot2)
```{r}
library(ggplot2)
x = ggplot(df2, aes(PPLCount))
x + geom_histogram()
```
     Here is a simple CDF for the PPLCount Variable
```{r}
p = ecdf(df2$PPLCount)
cdfAge = plot(p, ylab="Fn(Count of PPL)", xlab="Count of PPL", main="PPL Count (cdf)")
```

    Here is a more sophisticated CDF for the PPLCount Variable
```{r}
library(ggplot2)
ggplot(df2, aes(x = df2$PPLCount),) + stat_bin(aes(y = cumsum(..count..)), geom = "step", colour = "black", fill = "white", binwidth = 5) + theme_bw()
```

Question 5 Part 2
Dataset 2:

  Here is a summary of the data set
```{r}
summary(df2)
```
  Here is a summary of the PPLCount variable
```{r}
summary(df2$PPLCount)
```
  Here is a boxplot of the PPLCount variable
```{r}
boxplot(df2$PPLCount, main="Boxplot of People Count", ylab="Count", ylim=c(0,40))
```
  Here is a boxplot using ggplot2 with Data Points for the PPLCount variable
```{r}
ggplot(df2, aes(x = factor(0), y = PPLCount)) + geom_boxplot(width = .2) + ylab("Count of People") + xlab("") + geom_point(aes(x = factor(0), y = PPLCount), size = 1, colour = "blue") + theme_bw()
```

Question 5 Part 3
Dataset 2: 

    Here is the MLE for the PPLCount Variable
```{r}
theta = c(0,1)
fn = function(theta) { sum( 0.5*(df2$PPLCount-theta[1])^2/theta[2] + 0.5* log(theta[2]) ) }
nlm(fn, c(0,2), hessian = TRUE)
```
  I am not that sure on this part but from what I read online it looks as if the
  estimate of 22.49 is very close to my actual mean of: 
```{r}
mean(df2$PPLCount)
```
Question 5 Part 4
Dataset 2:

 After conducting the experiment and looking at my data, I realized that it would have been
 more valuable to have broken up each hour into smaller intervals.  Its important for me to 
 mention that there were a total of 4 registers.  3 registers were in my line of sight from 
 where I was sitting.  Therefore only three of the 4 total registers were included in the 
 observations.
  
 At the time of data collection, I was interested in seeing if there would be a great 
 difference between the number of people going to one particular register over another.  And 
 for this reason, I didnt pay much attention to the time interval part.  From the data you can 
 see that the number of people that checked out of registers 1 and 2 are about equal when you 
 take the total per hour count.  My hypothesis was that the 2nd register would have 
 more people checked out because the view of the 1st register was partialy blocked from the 
 customers line of sight by a 1/2 foot wide column.  What I ended up observing was that as the 
 number of people in one checkout line grew past 3, the next person in the line would stop to 
 see if any other registers were open.  This behavior is what allowed each register to 
 have roughly the same number of people checkout over the course of the 2 hour observation.  
   
#-----------------------------------------------------------------------------------------

Question 5 Part 1
Dataset 3:
    We first set up the working directory, grab a subset of the original dataset
    and rename columns.
```{r}
  setwd("~/Desktop/Problem Set 1")
  ds3 = "~/Desktop/Problem Set 1/random_variable3.tsv"
  df3 = read.csv(ds3, header = T, sep = ",")
```
  
    Here's a histogram representing the PMF for the PPLCount Variable
```{r, echo=FALSE}
hist(df3$PPLCount, prob=T, ylim=c(0, 0.25), breaks = 12, main = "People Count (pmf)", xlab = "Number of People", las = 1)
lines(density(df3$PPLCount), col=2, lwd=3)
```
    Heres another histogram representing the frequency distribution for the PPLCount   
    Variable (using ggplot2)
```{r}
library(ggplot2)
x = ggplot(df3, aes(PPLCount))
x + geom_histogram()
```

    Here is a simple CDF for the PPLCount Variable
```{r}
p = ecdf(df3$PPLCount)
cdfAge = plot(p, ylab="Fn(Count of People)", xlab="PPLCount", main="People Count (cdf)")
```

    Here is a more sophisticated CDF for the PPLCount Variable
```{r}
library(ggplot2)
ggplot(df3, aes(x = df3$PPLCount),) + stat_bin(aes(y = cumsum(..count..)), geom = "step", colour = "black", fill = "white", binwidth = 5) + theme_bw()
```

Question 5 Part 2
Dataset 3:

  Here is a summary of the data set
```{r}
summary(df3)
```
  Here is a summary of the PPLCount variable
```{r}
summary(df3$PPLCount)
```
  Here is a boxplot of the PPLCount variable
```{r}
boxplot(df3$PPLCount, main="Boxplot of People Count", ylab="Count", ylim=c(0,10))
```
  Here is a boxplot using ggplot2 with Data Points for the PPLCount variable
```{r}
ggplot(df3, aes(x = factor(0), y = PPLCount)) + geom_boxplot(width = .2) + ylab("Count") + xlab("") + geom_point(aes(x = factor(0), y = PPLCount), size = 1, colour = "blue") + theme_bw()
```

Question 5 Part 3
Dataset 3: 

    Here is the MLE for the PPLCount Variable
```{r}
theta = c(0,1)
fn = function(theta) { sum( 0.5*(df3$PPLCount-theta[1])^2/theta[2] + 0.5* log(theta[2]) ) }
nlm(fn, c(0,2), hessian = TRUE)
```
  I am not that sure on this part but from what I read online it looks as if the
  estimate of 4.49 is very close to my actual mean of: 
```{r}
mean(df3$PPLCount)
```
Question 5 Part 4
Dataset 3:

  I expected to see a normal distribution and that is what I saw after running the Histogram.  
  When collecting the data, I was kind of surprised by the number of 5 minute intervals where 
  there were exactly 4 people.  Learning from the 2nd dataset, I made sure that I collected 
  data in 5 minute intervals.  Doing this helped give me a better idea of the flow of people 
  per time 5 minute interval.  Even though this doesnt change the mean of the variable over 
  the course of the 2 hours, it gives you a more granular look at the change per every 5 
  minutes.  From this point you can perform calculations that tell you the rate of change from 
  one 5 minute interval to the next.  
